---
title: "Reaction Time Statistical Modeling"
date: "2024-04-27"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Project Overview

This project studies how demographic, behavioral, device, and environmental factors are associated with reaction time. The workflow includes data cleaning, exploratory data analysis, nonparametric testing, multiple linear regression, model selection, diagnostics, and a Box-Cox transformation.

The analysis is observational, so reported relationships should be interpreted as associations rather than causal effects.

# Packages and Data

```{r}
library(tidyverse)
library(corrplot)
library(car)
library(leaps)
library(lmtest)
```

```{r}
survey <- read.csv("Survey.csv")

glimpse(survey)
summary(survey$Reaction.time)
```

# Data Preparation

```{r}
survey <- survey %>%
  dplyr::mutate(
    Inputdevice.new = case_when(
      Input.device %in% c("Game controller", "Mouse", "Keyboard") ~ "Click-based",
      Input.device %in% c("Touch screen", "Trackpad") ~ "Tap-based",
      TRUE ~ NA_character_
    ),
    Inputdevice.new = factor(Inputdevice.new, levels = c("Click-based", "Tap-based")),

    Fatigue.new = case_when(
      Fatigue.level %in% c("Extremely fatigued", "Very fatigued") ~ "High",
      Fatigue.level == "Moderately fatigued" ~ "Moderate",
      Fatigue.level %in% c("Not fatigued at all", "Slightly Fatigued") ~ "Low",
      TRUE ~ NA_character_
    ),
    Fatigue.new = factor(Fatigue.new, levels = c("Low", "Moderate", "High")),

    Noise.new = case_when(
      as.character(Noise.level) %in% c("7", "8", "9") ~ "High",
      as.character(Noise.level) %in% c("4", "5", "6") ~ "Moderate",
      as.character(Noise.level) %in% c("1", "2", "3") ~ "Low",
      TRUE ~ NA_character_
    ),
    Noise.new = factor(Noise.new, levels = c("Low", "Moderate", "High")),

    RightHand.new = factor(
      ifelse(Primary.hand == "Right hand", "Yes", "No"),
      levels = c("Yes", "No")
    ),

    Visual.new = case_when(
      Visual.acuity %in% c("Very Poor", "Poor") ~ "Poor",
      Visual.acuity == "Average" ~ "Average",
      Visual.acuity == "Good" ~ "Good",
      Visual.acuity == "Excellent" ~ "Excellent",
      TRUE ~ NA_character_
    ),
    Visual.new = factor(Visual.new, levels = c("Excellent", "Good", "Average", "Poor")),

    Temp.new = case_when(
      Temp.level %in% c("Very Warm", "Warm") ~ "Warm",
      Temp.level %in% c("Very Cold", "Cold") ~ "Cold",
      Temp.level == "Neutral" ~ "Neutral",
      TRUE ~ NA_character_
    ),
    Temp.new = factor(Temp.new, levels = c("Neutral", "Cold", "Warm")),

    Cautious.new = case_when(
      Cautious.level %in% c("Extremely cautious", "Very cautious") ~ "High",
      Cautious.level == "Moderately cautious" ~ "Moderate",
      Cautious.level %in% c("Not cautious at all", "Slightly cautious") ~ "Low",
      TRUE ~ NA_character_
    ),
    Cautious.new = factor(Cautious.new, levels = c("Low", "Moderate", "High")),

    is_apple = factor(ifelse(
      grepl("macOS|iOS|IOS|Apple|iPhone|iPad", Device.OS),
      "Apple", "Non-Apple"
    )),
    is_mobile = factor(ifelse(
      grepl("Smartphone|Tablet|iPhone|iPad", Device.OS),
      "Mobile", "Non-Mobile"
    ))
  ) %>%
  dplyr::select(-Alcohol.intake)
```

# Exploratory Data Analysis

## Reaction-Time Distribution

```{r}
ggplot(survey, aes(x = Reaction.time)) +
  geom_histogram(bins = 25, boundary = 0) +
  labs(
    title = "Distribution of Reaction Time",
    x = "Reaction time (ms)",
    y = "Count"
  ) +
  theme_minimal()
```

## Correlation Among Numeric Variables

```{r}
numerical_data <- survey %>% dplyr::select(where(is.numeric))
corr_matrix <- cor(numerical_data, use = "complete.obs")
corrplot(corr_matrix, method = "circle")
```

## Selected Bivariate Relationships

```{r}
ggplot(survey, aes(x = Age, y = Reaction.time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Reaction Time vs. Age", x = "Age", y = "Reaction time (ms)") +
  theme_minimal()

ggplot(survey, aes(x = Avg.sleep.time, y = Reaction.time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Reaction Time vs. Average Sleep", x = "Average sleep (hours)", y = "Reaction time (ms)") +
  theme_minimal()

ggplot(survey, aes(x = Inputdevice.new, y = Reaction.time)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.12) +
  labs(title = "Reaction Time by Input Device", x = "Input-device group", y = "Reaction time (ms)") +
  theme_minimal()
```

# Input-Device Comparison

```{r}
click_based <- survey$Reaction.time[survey$Inputdevice.new == "Click-based"]
tap_based <- survey$Reaction.time[survey$Inputdevice.new == "Tap-based"]

shapiro.test(click_based)
shapiro.test(tap_based)
wilcox.test(click_based, tap_based, alternative = "two.sided")
```

The original analysis found evidence that reaction-time distributions differed between click-based and tap-based input-device groups.

# Focused Regression: Age and Input Device

```{r}
model_age <- lm(Reaction.time ~ Age, data = survey)
model_device <- lm(Reaction.time ~ Inputdevice.new, data = survey)
model_additive <- lm(Reaction.time ~ Age + Inputdevice.new, data = survey)
model_interaction <- lm(Reaction.time ~ Age * Inputdevice.new, data = survey)

summary(model_additive)
summary(model_interaction)

anova(model_age, model_additive)
anova(model_additive, model_interaction)
```

The interaction model tests whether the age slope differs across input-device groups.

```{r}
ggplot(survey, aes(x = Age, y = Reaction.time, color = Inputdevice.new)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(
    title = "Age and Reaction Time by Input-Device Group",
    x = "Age",
    y = "Reaction time (ms)",
    color = "Input device"
  ) +
  theme_minimal()
```

# Multiple Regression and Model Selection

```{r}
survey_model <- survey %>%
  dplyr::select(
    Reaction.time, Class, Avg.sleep.time, last.night.sleep.time, Awake.hours,
    Stress.level, Distraction, Noise.new, Temp.new, Game.freq, Sport.freq,
    Avg.hours.exercise, Caffein.intake, is_mobile, is_apple, Age,
    Cautious.new, Inputdevice.new, Visual.new, RightHand.new,
    Fatigue.new, WiFi.stable, Use.primary.hand
  ) %>%
  tidyr::drop_na()

full_model <- lm(Reaction.time ~ ., data = survey_model)
summary(full_model)
```

Best-subset regression is used as a screening tool, with adjusted R-squared, Mallows' Cp, AIC, and BIC used to compare candidate model sizes.

```{r}
subset_fit <- regsubsets(Reaction.time ~ ., data = survey_model, nvmax = 12)
subset_summary <- summary(subset_fit)

n <- nrow(survey_model)
model_size <- seq_along(subset_summary$rss) + 1
AIC_values <- n * log(subset_summary$rss / n) + 2 * model_size
BIC_values <- n * log(subset_summary$rss / n) + model_size * log(n)

par(mfrow = c(2, 2))
plot(model_size, subset_summary$adjr2, xlab = "Number of parameters", ylab = "Adjusted R-squared")
plot(model_size, subset_summary$cp, xlab = "Number of parameters", ylab = "Mallows' Cp")
plot(model_size, AIC_values, xlab = "Number of parameters", ylab = "AIC")
plot(model_size, BIC_values, xlab = "Number of parameters", ylab = "BIC")
par(mfrow = c(1, 1))
```

```{r}
reduced_model <- lm(
  Reaction.time ~ WiFi.stable + Visual.new + Inputdevice.new +
    Avg.hours.exercise + Sport.freq + Temp.new + Stress.level + Class,
  data = survey_model
)

summary(reduced_model)
anova(reduced_model, full_model)
```

# Model Diagnostics

```{r}
par(mfrow = c(2, 2))
plot(reduced_model)
par(mfrow = c(1, 1))

car::vif(reduced_model)
lmtest::bptest(reduced_model)
shapiro.test(resid(reduced_model))
car::durbinWatsonTest(reduced_model)
```

For the Breusch-Pagan test, the null hypothesis is constant variance. A small p-value is evidence of heteroskedasticity. For the Shapiro-Wilk test, a small p-value is evidence against residual normality. The Durbin-Watson test is used to assess residual autocorrelation.

# Box-Cox Transformation

```{r}
MASS::boxcox(reduced_model, plotit = TRUE)

lambda <- -0.5
survey_model <- survey_model %>%
  dplyr::mutate(
    Transformed_Reaction.time = (Reaction.time^lambda - 1) / lambda
  )

transformed_model <- lm(
  Transformed_Reaction.time ~ WiFi.stable + Visual.new + Inputdevice.new +
    Avg.hours.exercise + Sport.freq + Temp.new + Stress.level + Class,
  data = survey_model
)

summary(transformed_model)
```

```{r}
par(mfrow = c(2, 2))
plot(transformed_model)
par(mfrow = c(1, 1))

lmtest::bptest(transformed_model)
shapiro.test(resid(transformed_model))
car::durbinWatsonTest(transformed_model)
```

# Influence Diagnostics

```{r}
n <- nrow(survey_model)
p <- length(coef(transformed_model))

leverage <- hatvalues(transformed_model)
high_leverage <- which(leverage > 2 * p / n)
high_leverage

studentized <- rstudent(transformed_model)
bonferroni_cutoff <- abs(qt(0.05 / (2 * n), df.residual(transformed_model)))
which(abs(studentized) > bonferroni_cutoff)

cooks_d <- cooks.distance(transformed_model)
which(cooks_d >= 1)

plot(leverage, main = "Leverage Values", ylab = "Leverage")
abline(h = 2 * p / n, lty = 2)

plot(studentized, main = "Studentized Residuals", ylab = "Studentized residual")
abline(h = c(-bonferroni_cutoff, bonferroni_cutoff), lty = 2)

plot(cooks_d, main = "Cook's Distance", ylab = "Cook's distance")
abline(h = 1, lty = 2)
```

# Nonlinearity Check for Exercise Hours

```{r}
transformed_quadratic <- lm(
  Transformed_Reaction.time ~ WiFi.stable + Visual.new + Inputdevice.new +
    Avg.hours.exercise + I(Avg.hours.exercise^2) + Sport.freq +
    Temp.new + Stress.level + Class,
  data = survey_model
)

anova(transformed_model, transformed_quadratic)
```

# Takeaways

This project demonstrates an end-to-end statistical-modeling workflow: feature engineering, exploratory analysis, nonparametric testing, nested-model comparison, subset selection, regression diagnostics, response transformation, and influence analysis.

Because the data are observational and several predictors are self-reported, results should be interpreted as associations rather than causal effects.
